pirouette: the error BEAST2 makes in inferring a phylogeny

### 2019-01-21 article.tex
Methods first draft done

An optional step is to generate a 'twin tree', that will be
analyzed in the same way as the true tree.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Usage}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

A first research question that \verb;pirouette; answers, is:

What is the error BEAST2 makes from a phylogeny using the same 
speciation model as it was generated by?

Say we have an idealised

\begin{lstlisting}[language=R, floatplacement=H]
phylogeny <- ape::read.tree(text = "((A:4, B:4):1, (C:4, D:4):1);")
\end{lstlisting}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{pirouette resources}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

An optional step is to generate a 'twin tree', that will be
analyzed in the same way as the true tree. The twin tree has the same
topology as the given phylogeny, yet its branch lengths follow the same
distribution as a Yule (pure-birth) or (constant rate) birth-death tree prior.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Usage: first research question}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

A first research question that \verb;pirouette; answers, is:
What is the error BEAST2 makes from a phylogeny using the same 
speciation model as it was generated by?

We start with an idealised birth-death tree of five taxa and a crown age of 10:

% Need idealised birth-death tree
% From https://github.com/richelbilderbeek/pirouette_article/issues/1
% and https://github.com/richelbilderbeek/pirouette/issues/35
\begin{lstlisting}[language=R, floatplacement=H]
phylogeny <- ape::read.tree(text = "((A:4, B:4):1, (C:4, D:4):1);")
\end{lstlisting}

The first step in \verb;pirouette; is to simulate a DNA alignment from the 
given phylogeny. To do so, we must specify the DNA root sequence
and a mutation rate. In this example, the DNA root sequence consists
out of four block of 250 nucleotides each, where the per-nucleotide
mutation rate is on average 0.1 mutations per time unit:

\begin{lstlisting}[language=R, floatplacement=H]
alignment_params <- create_alignment_params(
  root_sequence = create_blocked_dna(length = 1000),
  mutation_rate = 0.1
)
\end{lstlisting}

The second step is to selects how an inference model is picked.
We just select the generative model, which are the site and clock
model of the alignment. Additionally, we specify a birth-death tree prior
to be the generative tree prior.

% Note that this syntax does not work yet, see
% https://github.com/richelbilderbeek/pirouette/issues/36
\begin{lstlisting}[language=R, floatplacement=H]
model_select_param <- create_gen_model_select_param(
  alignment_params = alignment_params,
  tree_prior = create_bd_tree_prior()
)
\end{lstlisting}

The third and fourth step have sensible defaults, and we are not
interested in using a twin tree. We can now measure the errors BEAST2
makes when inferring the given phylogeny like this:

\begin{lstlisting}[language=R, floatplacement=H]
errors <- pir_run(
  phylogeny,
  alignment_params = alignment_params,
  model_select_params = model_select_param
)
\end{lstlisting}

\verb;pirouette; has a nice plotting function:

\begin{lstlisting}[language=R, floatplacement=H]
pir_plot(errors)
\end{lstlisting}

% TODO: write script to create figure 2, see
% https://github.com/richelbilderbeek/pirouette/issues/37
The resulting figure is shown in figure 2.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Usage: second research question}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

A second research question that \verb;pirouette; answers, is:
What is the error BEAST2 makes from a phylogeny 
picking the best inference model?

Here we start with an tree from an unknown speciation model,
that has four taxa and a crown age of 5:

% Need idealised birth-death tree
% From https://github.com/richelbilderbeek/pirouette_article/issues/1
% and https://github.com/richelbilderbeek/pirouette/issues/35
\begin{lstlisting}[language=R, floatplacement=H]
phylogeny <- ape::read.tree(text = "((A:4, B:4):1, (C:4, D:4):1);")
\end{lstlisting}

The first step in \verb;pirouette; is to simulate a DNA alignment from the 
given phylogeny. We'll use the same parameters here:

\begin{lstlisting}[language=R, floatplacement=H]
alignment_params <- create_alignment_params(
  root_sequence = create_blocked_dna(length = 1000),
  mutation_rate = 0.1
)
\end{lstlisting}

The second step is to select how an inference model is picked.
We will let the inference model that has the highest evidence be used
in the Bayesian inference. By default, \verb;pirouette; will try
all combinations of site models, clock models and tree priors.

% Note that this syntax does not work yet, see
% https://github.com/richelbilderbeek/pirouette/issues/36
\begin{lstlisting}[language=R, floatplacement=H]
model_select_param <- create_best_model_select_param()
\end{lstlisting}

Also here, the third and fourth step have sensible defaults, and we are not
interested in using a twin tree. We can now measure the errors BEAST2
makes when inferring the given phylogeny like this:

\begin{lstlisting}[language=R, floatplacement=H]
errors <- pir_run(
  phylogeny,
  alignment_params = alignment_params,
  model_select_params = model_select_param
)
\end{lstlisting}

Again, showing the results:

\begin{lstlisting}[language=R, floatplacement=H]
pir_plot(errors)
\end{lstlisting}

% TODO: write script to create figure 3, see
% https://github.com/richelbilderbeek/pirouette/issues/38
The resulting figure is shown in figure 3.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Usage: third research question}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

A third research question that \verb;pirouette; answers, is:
What is the error BEAST2 makes from a phylogeny 
picking the best inference model, compared to the background noise?

Here we again start with an tree from an unknown speciation model,
that has four taxa and a crown age of 5:

% Need idealised birth-death tree
% From https://github.com/richelbilderbeek/pirouette_article/issues/1
% and https://github.com/richelbilderbeek/pirouette/issues/35
\begin{lstlisting}[language=R, floatplacement=H]
phylogeny <- ape::read.tree(text = "((A:4, B:4):1, (C:4, D:4):1);")
\end{lstlisting}

The first step in \verb;pirouette; is to simulate a DNA alignment from the 
given phylogeny. We'll use the same parameters here:

\begin{lstlisting}[language=R, floatplacement=H]
alignment_params <- create_alignment_params(
  root_sequence = create_blocked_dna(length = 1000),
  mutation_rate = 0.1
)
\end{lstlisting}

The second step is to select how an inference model is picked.
We will let the inference model that has the highest evidence be used
in the Bayesian inference. By default, \verb;pirouette; will try
all combinations of site models, clock models and tree priors.

% Note that this syntax does not work yet, see
% https://github.com/richelbilderbeek/pirouette/issues/36
\begin{lstlisting}[language=R, floatplacement=H]
model_select_param <- create_best_model_select_param()
\end{lstlisting}

Also here, the third and fourth step have sensible defaults.

This time, we are interested in creating a twin tree. A twin tree
has the same topology as the given tree, yet its branch lengths follow
a birth-death distribution. Creating a twinning parameter is easy,
as it has sensible default settings:

\begin{lstlisting}[language=R, floatplacement=H]
twinning_params <- create_twinning_params()
\end{lstlisting}

We can now measure the errors BEAST2
makes when inferring the given phylogeny, compared
to the background noise, like this:

\begin{lstlisting}[language=R, floatplacement=H]
errors <- pir_run(
  phylogeny,
  twinning_params = twinning_params,
  alignment_params = alignment_params,
  model_select_params = model_select_param
)
\end{lstlisting}

Again, showing the results:

\begin{lstlisting}[language=R, floatplacement=H]
pir_plot(errors)
\end{lstlisting}

% TODO: write script to create figure 4, see
% https://github.com/richelbilderbeek/pirouette/issues/39
The resulting figure is shown in figure 4
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{pirouette resources}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
